---
title: 'gganimate: Animations with ggplot2'
author: Thean Lim
date: '2018-04-19'
slug: gganimate-animations-with-ggplot2
categories:
  - R
tags:
  - visualization
  - animation
header:
  caption: ''
  image: ''
summary: How to create animated plots from ggplot2?
output:
  blogdown::html_page:
    toc: true
    number_sections: false
    toc_depth: 3
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(message = FALSE, warning = FALSE)
options(width = 100, dplyr.width = 100)
library(methods)
library(ggplot2)
theme_set(theme_bw())
```
The following content is based on https://github.com/dgrtwo/gganimate

# Install
```{r}
if(!require(devtools)) install.packages("devtools")
if(!require(gganimate)) devtools::install_github("dgrtwo/gganimate")
```

The main argument is the `frame` ( the time frame/ time points in the argument). Treat `frame` as another aesthetics such as **size, color, x and y**. Then, use the `gganimate()` function.

# Demo dataset

The dataset that will be used is `gapminder`. This is an excerpt of the Gapminder data on life expectancy, GDP per capita, and population by country. Let's install the package to get the data and take a look on it.

```{r}
if(!require(gapminder)) install.packages("gapminder")
# Loading the two libraries required
library(gapminder)
head(gapminder)
```

## Simple Animation

- Build a scatterplot the Life Expectancy against gdp per capita across years.
- Use color and sizes to show the continents and population size.

** The only thing different from usual plotting is the `frame` aesthetics. Then, we use the `gganimate()` function**.

```{r,fig.show = "animate", ffmpeg.format='gif', dev='jpeg'}
# Load required package
library(gapminder)
library(ggplot2)
library(gganimate)

## Basic Scatterplot
p <- ggplot(gapminder, mapping = aes(x = gdpPercap, y = lifeExp, 
                           size = pop, color = continent, frame = year)) +
  geom_point() +
  scale_x_log10()

# Animation
gganimate(p)
```
Note that the axis remains the same, making comparison across time easier.

{{% alert note %}}
Set `fig.show = "animate"` chunk option if you are using Rmd and knitr.
Add in `fig.show = "animate", ffmpeg.format='gif', dev='jpeg'` if you are using blogdown.
{{% /alert %}}

The animation could be saved by the followings:

```{r, eval = FALSE}
gganimate(p, "output.gif")
gganimate(p, "output.mp4")
gganimate(p, "output.swf")
gganimate(p, "output.html")
```

{{% alert note %}}
ffmpeg and ImageMagick drivers should be installed on your computer.
{{% /alert %}}

{{% alert warning %}}
I met some problems with running the `gganimate(p)` and I found the following solution.

[Solutions to "Installing #ImageMagick for #gganimate R package" errors](https://medium.com/@DaveQuartey/solutions-to-installing-imagemagick-for-gganimate-r-package-errors-8b36d0736f85)
{{% /alert %}}

## Fixing running `gganimate()` problem

Just in case that the page is removed in the future, I am going to copy and paste whatever is in that page:

### Install ImageMagick through R

Do this through the following. Make sure to select `legacy utilities` whiles installing ImageMagick to install convert.exe on your machine.

```{r, eval = FALSE}
if (!require("installr")){ 
install.packages("installr")
library(installr)
install.imagemagick("https://www.imagemagick.org/script/download.php"
} else {
library(installr)
install.imagemagick("https://www.imagemagick.org/script/download.php")
```

This downloads ImageMagick onto the PC. Rerun the `gganimate()` code after this step.

### Error after installing ImageMagick

```
cannot find ImageMagick with convert = "convert"
```
Install ImageMagick and point R path to its folder . Make sure to restart RStudio after pointing R path to its folder.

#### Path through R

[Point](https://github.com/dgrtwo/gganimate/issues/22#issuecomment-264352060) R path to its folder in R using:

```
Sys.setenv(PATH = paste("C:/Program Files/ImageMagick/bin",
Sys.getenv("PATH"), sep = ";"))
```

#### Restart Rstudio

if the following error comes up after installing ImageMagick and directing R to its path. Just restart Rstudio.

```
C:\Program' is not recognized as an internal or external command,
operable program or batch file.
```
## Customization

### Highlight points (rather than moving them)

What if I want to highlight the change in the `population` throughout the years but not moving it?

All I need to do is to remove the `frame` aesthetics from the global chunk but add it into specific `geom`.

```{r, fig.show = "animate",ffmpeg.format='gif'}
p2 <- ggplot(gapminder, mapping = aes(x = gdpPercap, y = lifeExp, size = pop)) +
  geom_point() +
  geom_point(aes(frame = year), color = "violetred") +
  scale_x_log10()

gganimate(p2)
```

### Cumulative Layer

This can be done by adding an aesthetic `cumulative = TRUE`. It's useful for building path over time. Also note that the `group` aesthetics is used so that it knows which variables to accumulate on.

```{r, fig.show = "animate",ffmpeg.format='gif'}
p3 <- ggplot(gapminder, mapping = aes(x = gdpPercap, y = lifeExp, frame = year)) +
  geom_path(aes(cumulative = TRUE, group = country)) +
  scale_x_log10() +
  facet_wrap(~continent)

gganimate(p3)
```

### Matching `frame` with other dimension (variables)

Let's try to match `continent` to `frame` aesthetics

```{r, fig.show = "animate",ffmpeg.format='gif' }
p4 <- ggplot(gapminder, 
             mapping = aes(x = gdpPercap, y = lifeExp, size = pop, frame = continent)) +
  geom_point(color = "blue") +
  scale_x_log10()

gganimate(p4)
```

### Statistics Summarization

Always add in `group` in the `geom` (such as `geom_smoooth()`) or the statistics will be calculated across all frames and thus be constant in animation.  

`scales = "free"` was added because the range of x axis is fixed by default.  

```{r, fig.show = "animate",ffmpeg.format='gif'}
p5 <- ggplot(gapminder, 
             mapping = aes(x = gdpPercap, y = lifeExp, size = pop, frame = year)) +
  geom_point() +
  geom_smooth(aes(group = year), method = "lm", show.legend = FALSE) +
  facet_wrap(~continent, scales = "free") +
  scale_x_log10() 

gganimate(p5)
```


### Animation Options

You can control your animation with the same options described in `ani.options` from the `animation` package, passed directly to `gganimate` or to `ani.options` beforehand. 
Note that you always need to add the parameters passed to `gganimate` to the chunks in the knitr of Rmd.

Now, let's try to set the `interval` argument and make the animation goes faster.  

We will start with passing the argument directly intto `gganimate()`

```{r, fig.show = "animate",ffmpeg.format='gif', interval=.2}
gganimate(p, interval = .2)
```

In fact, simply making the chunk to include `interval = 0.2` is good enough in Rmd.

```{r, fig.show = "animate",ffmpeg.format='gif', interval = .2}
gganimate(p)
```

Let's do the same thing but a instead using the `ani.options`. This is how you do it in Rstudio in general.

```{r, fig.show = "animate",ffmpeg.format='gif', interval = .2}
library(animation)
ani.options(interval = .2)
gganimate(p)
```



